---
title: "Hampshire GHG emissions (CO2e analysis)"
author: "Ben Anderson"
format: html
self-contained: true
editor: visual
---

```{r}
#| label: setup
library(data.table)
library(flextable)
library(ggplot2)
library(here)

source(here::here("R", "functions.R"))
qmdParams <- list()

qmdParams$dataPath <- "~/Dropbox/data/beis/localAuthority/allGHG/2005-21-local-authority-ghg-emissions-csv-dataset-update-060723.csv" # includes all GHGs
```

# Introduction

Latest CO2e data from BEIS at [district](https://www.gov.uk/government/statistics/uk-local-authority-and-regional-greenhouse-gas-emissions-national-statistics-2005-to-2020) level - now has CO2e (i.e methane etc as well)

# Load and filter data

```{r}
#| label: loadData
dt <- data.table::fread(qmdParams$dataPath)
```

How many years are available?

```{r}
#| label: checkYears
summary(dt$`Calendar Year`)
```

Extract just the 14 districts we are interested in.

```{r}
#| label: filter_areas
dt[, la_name := `Local Authority`]

maxYear <- max(dt$`Calendar Year`)

solent <- getSolent(dt)

t <- solent[`Calendar Year` == maxYear & `Greenhouse gas` == "CO2", .(n = .N,
                           sumkT_CO2e = sum(`Territorial emissions (kt CO2e)`)), keyby = .(`Local Authority`)]

makeFlexTable(t[order(-sumkT_CO2e)], cap = paste0("Check LAs (sum CO2 for ", maxYear, ", kt CO2e"))

plotDT <- solent[, .(n = .N,
                           sumkT_CO2e = sum(`Territorial emissions (kt CO2e)`)), keyby = .(`Local Authority`, `Greenhouse gas`, `Calendar Year`)]

ggplot2::ggplot(plotDT[`Calendar Year` > 2015], aes(x = `Calendar Year`, y = sumkT_CO2e,
                            fill = `Local Authority`)) +
  geom_col(position = "stack") +
  facet_grid(`Greenhouse gas` ~ .) +
  labs(y = "kT CO2e")
```

# Analysis

## Land Use: Sequestration estimates

The land use data includes the categories shown in Table XX. Of these we would expect net positive emissions from cropland and settlements and also net positive emissions from CH4 and N2O. To understand the potential for sequestration we will therefore need to filter out these two categories and these two GHGs.

```{r}
#| label: extractLandUse
# Net Emissions: Forest land	
# Net Emissions: Cropland	
# Net Emissions: Grassland	
# Net Emissions: Wetlands	
# Net Emissions: Settlements	
# Net Emissions: Harvested Wood Products	
# Net Emissions: Indirect N2O
solent_LULUCF <- solent[`LA GHG Sub-sector` %like% "LULUCF"]

t <- solent_LULUCF[, .(nYears = uniqueN(`Calendar Year`),
                       min_kTCO2e = min(`Territorial emissions (kt CO2e)`),
                       mean_kTCO2e = mean(`Territorial emissions (kt CO2e)`),
                       max_kTCO2e = max(`Territorial emissions (kt CO2e)`)),
                   keyby = .(`Greenhouse gas`,`LA GHG Sub-sector`)]
makeFlexTable(t, cap = "Min, mean and max values for LULUCF sub-sectors by GHG")
```

Figure @landUsePlot shows...

```{r}
#| label: landUsePlot
#| fig-cap: "Land use plot - all GHGs by sector over time"
#| fig-link: fig-landUsePlot

ggplot2::ggplot(solent_LULUCF, aes(x = `Calendar Year`,
                                   y = `Territorial emissions (kt CO2e)`,
                                   colour = `Local Authority`)) +
  geom_line() +
  facet_grid(`LA GHG Sub-sector` ~ `Greenhouse gas`) +
  theme(legend.position = "bottom")

```

Selecting just CO2 data for:

-   Forest land

-   Grassland

-   Wetlands

```{r}
#| label: sequestrationTotals

t <- solent_LULUCF[`LA GHG Sub-sector` %like% "Forest" |
                     `LA GHG Sub-sector` %like% "Grassland" |
                     `LA GHG Sub-sector` %like% "Wetland", .(sum_kT_Co2e = sum(`Territorial emissions (kt CO2e)`)), 
                   keyby =.(la_name = `Local Authority`, 
                            year = `Calendar Year`)]

t_dcast <- data.table::dcast(t[year > 2017], 
                             la_name ~ year)

makeFlexTable(t_dcast, cap = "Land use 'sequestration' sum by district (kT CO2e)")

annualtotals <- t[year > 2017, .(kT_CO2e = sum(sum_kT_Co2e)), keyby = .(year)]
annualtotals[, year := as.factor(year)]
makeFlexTable(annualtotals, cap = "Land use 'sequestration' sum (kT CO2e)")
message("Districts included:")
t[, .(n = .N), keyby = .(la_name)]
```

```{r}
#| label: theEnd
```
